---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2023"
subtitle: "Assignment 4 - Due date 02/17/23"
author: "Wendy Wen"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: inline
---
  
## Directions
  
You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. And to do so you will need to fork our repository and link it to your RStudio.  

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A04_Sp23.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

R packages needed for this assignment: "xlsx" or "readxl", "ggplot2", "forecast","tseries", and "Kendall". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r}
#Load/install required package here
library(lubridate)
library(forecast)
library(tseries)
library(Kendall)
library(readxl)
library(ggplot2)
library(tidyr)
```

## Questions 

Consider the same data you used for A3 from the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx". The data comes from the US Energy Information and Administration and corresponds to the December 2022 Monthly Energy Review. For this assignment you will work only with the column "Total Renewable Energy Production". 


```{r}
#Importing data set - using xlsx package
energy_data <- read_excel(path="/Users/wendywen/Desktop/ENV790/wendywen_TSA/Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",skip = 12, sheet="Monthly Data",col_names=FALSE) 

#Now let's extract the column names from row 11
read_col_names <- read_excel(path="/Users/wendywen/Desktop/ENV790/wendywen_TSA/Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",skip = 10,n_max = 1, sheet="Monthly Data",col_names=FALSE) 
colnames(energy_data) <- read_col_names
date <- energy_data[,1]

data4 <- energy_data[,c("Total Renewable Energy Production")]
col_names <- colnames(data4)
nobs <- nrow(data4)
df_data4 <- as.data.frame(data4)

ts_data4 <- ts(df_data4, start = c(1973,1), frequency=12)
```


## Stochastic Trend and Stationarity Tests

### Q1

Difference the "Total Renewable Energy Production" series using function diff(). Function diff() is from package base and take three main arguments:
* *x* vector containing values to be differenced;
* *lag* integer indicating with lag to use;
* *differences* integer indicating how many times series should be differenced.

Try differencing at lag 1 only once, i.e., make `lag=1` and `differences=1`. Plot the differenced series. Do the series still seem to have trend?

```{r}
dif <- diff(ts_data4[,1], lag = 1, differences = 1)
dif <- as.data.frame(dif)
plot(dif[,1],type="l",col="blue",ylab= paste0(col_names[1]," [Trillion Btu]"),
     main=paste0("Differenced ",col_names[1]))
```

The differenced series of total renewable energy production seems to have no trend.

### Q2

Now let's compare the differenced series with the detrended series you calculated on A3. In other words, for the "Total Renewable Energy Production" compare the differenced series from Q1 with the series you detrended in A3 using linear regression. (Hint: Just copy and paste part of your code for A3)

Copy and paste part of your code for A3 where you compute regression for Total Energy Production and the detrended Total Energy Production

```{r}
#Create vector t
t <- 1:nobs

#Fit a linear trend
linear <- summary(lm(ts_data4[,1] ~ t))
linear

beta0 = as.numeric(linear$coefficients[1])
beta1 = as.numeric(linear$coefficients[2])
detrend <- df_data4[,1]-(beta0 + beta1*t)

det <- as.data.frame(detrend)
plot(date, det[,1], type="l",col="orange", xlab = "Time", ylab= paste0(col_names[1]," [Trillion Btu]"),main=paste0("Detrended ",col_names[1]))
```


### Q3
Create a data frame with 4 columns: month, original series, detrended by Regression Series and differenced series. Make sure you properly name all columns. Also note that the differenced series will have only 584 rows because you loose the first observation when differencing. Therefore, you need to remove the first observations for the original series and the detrended by regression series to build the new data frame.

```{r}
#Data frame - remember to not include January 1973
df_ren <- data.frame(date[2:597,], df_data4[2:597,], det[2:597,], dif)
colnames(df_ren) <- c("Date", "Original", "Detrended", "Differenced")
col_ren <- colnames(df_ren[,2:4])
head(df_ren)
```

### Q4
Using ggplot() create a line plot that shows the three series together. Make sure you add a legend to the plot.

```{r}
#Use ggplot
cols <- c("Original" = "green", "Detrended" = "orange", "Differenced" = "blue")
plt <- ggplot(df_ren) + 
  geom_line(aes(x = Date, y = Original, color = "Original")) +
  geom_line(aes(x = Date, y = Detrended, color = "Detrended")) +
  geom_line(aes(x = Date, y = Differenced, color = "Differenced")) +
  xlab("Date") + 
  ylab("Energy Production [Trillion Btu]") +
  labs(title ="Different Time Series of Total Renewable Energy Production") + scale_color_manual(name = "Series", values = cols) +
  theme_bw()
plt
```


### Q5
Plot the ACF for the three series and compare the plots. Add the argument `ylim=c(-0.5,1)` to the Acf() function to make sure all three y axis have the same limits. Which method do you think was more efficient in eliminating the trend? The linear regression or differencing?

```{r}
#Compare ACFs
par(mfrow=c(1,3))
for (i in 1:3){
  Acf(df_ren[i+1],lag.max = 40, main = paste0("ACF of ",col_ren[i]," Series"), ylim = c(-0.5,1))
}
```
Differencing is more efficient in eliminating the trend. Because the ACF of the differenced series shows relatively random correlations while the ACF of the detrended series (by linear regression) still shows an overall decayed pattern which implies that there still exist a trend.

### Q6
Compute the Seasonal Mann-Kendall and ADF Test for the original "Total Renewable Energy Production" series. Ask R to print the results. Interpret the results for both test. Whats the conclusion from the Seasonal Mann Kendall test? What's the conclusion for the ADF test? Do they match what you observed in Q2? Recall that having a unit root means the series has a stochastic trend. And when a series has stochastic trend we need to use a different procedure to remove the trend.

```{r}
# Seasonal Mann Kendall test of the original series
SMKtest <- SeasonalMannKendall(ts_data4[,1])
print("Results for Seasonal Mann Kendall")
print(summary(SMKtest))
#p value < 0.05, reject the null hypothesis, data has a trend

# ADF test of the original series
print("Results for ADF test")
print(adf.test(ts_data4[,1],alternative = "stationary"))
#p value > 0.05, fail to reject the null hypothesis, data has a unit root, i.e., stochastic trend
```

The Kendall rank correlation coefficient is 0.727 and the corresponding two-sided p-value is smaller than 2.22e-16, and therefore, we will reject the null hypothesis and conclude that a trend is present in the data.
In the ADF test, the test statistic and p-value are -1.2055 and 0.9056 respectively. Since the p-value is greater than 0.05, we will fail to reject the null hypothesis. In other words, the time series is non-stationary and has some time-dependent structure.
The results match what I observed in Q2. The time series has a non-stationary trend, and thus we need to use a different procedure to remove the trend. Applying the linear regression model to detrend the series, some trends still present and some different trends were created, implying linear regression is not the most ideal model.

### Q7

Aggregate the original "Total Renewable Energy Production" series by year. You can use the same procedure we used in class. Store series in a matrix where rows represent months and columns represent years. And then take the columns mean using function colMeans(). Recall the goal is the remove the seasonal variation from the series to check for trend. 

```{r}
date2 <- as_date(energy_data$Month, tz = NULL)

ren_matrix <- matrix(ts_data4[,1],byrow=FALSE,nrow=12)
ren_yearly <- colMeans(ren_matrix)

library(dplyr)  #move this to package chunk later
year <- c(year(first(date2)):year(last(date2)))
df_ren_yearly <- data.frame(year, ren_yearly)
head(df_ren_yearly)
```

### Q8

Apply the Mann Kendal, Spearman correlation rank test and ADF. Are the results from the test in agreement with the test results for the non-aggregated series, i.e., results for Q6?

```{r}
# Mann Kendal
print("Results of Mann Kendall on average yearly series")
print(summary(MannKendall(ren_yearly)))

# Spearman correlation rank test (Deterministic trend)
print("Results from Spearman Correlation")
cor_sp=cor.test(ren_yearly,year,method="spearman")
print(cor_sp)

# ADF
print("Results for ADF test on yearly data")
print(adf.test(ren_yearly, alternative = "stationary"))
```

For the average yearly seires, the Kendall rank correlation coefficient is 0.739 and the corresponding two-sided p-value is smaller than 2.22e-16, and therefore, we will reject the null hypothesis and conclude that a trend is present in the data. This is in agreement with the test results for the non-aggregated series.
In the Spearman's rank correlation test, the test statistic is 0.877 and p-value is smaller than 2.22e-16. Since the p-value is smaller than 0.05, we will reject the null hypothesis and conclude that the series has a monotonic trend.
In the ADF test, the test statistic and p-value are -1.6251 and 0.7247 respectively. Since the p-value is greater than 0.05, we will fail to reject the null hypothesis. In other words, the time series is non-stationary and has some time-dependent structure. This is in agreement with the test results for the non-aggregated series.

